Performance of Individual Audit Filters in Predicting Opportunities for Improvement

### 2024-06-18 manuscript.Rmd
update

---
title: Performance of Individual Audit Filters in Predicting Opportunities for Improvement
  in Adult Trauma Patients
author: "Josef Al-Khalili"
output:
  html_document: default
  pdf_document: default
subtitle: "A single-center registry-based retrospective cohort"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options:
  markdown:
    wrap: 72
---
```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(pROC)
library(gtsummary)
library(gt)
library(boot)
library(webshot2)
library(knitr)
```
# Abstract

@@

### Morbidity and Mortality Conferences
The MMC is a central QI technique focusing on anonymity, critical analysis of specific adverse events, recognizing flawed approaches, and proposing and implementing improvements [@gregor2016morbidity]. Attended by healthcare professionals involved in trauma care, MMCs discuss themes like unexpected mortality, morbidity, and errors [@orlander2003morbidity], fostering error reduction and encouraging reporting without negative feedback [@lazzara2022morbidity]. Effective MMC implementation has been shown to increase morbidity and mortality reporting [@kong2016analysis].

@@

@@

## Setting
Karolinska University Hospital, a Trauma level I
hospital, treats around 1,500 trauma patients annually. If a patient case results in team activation or the
patient has an Injury Severity Score (ISS) of less than 9
retrospectively, it is added to the Karolinska
trauma registry which reports to SweTrau. The registry includes data on vital
signs, times, injuries, and interventions as well as patient
demographics according to the European Consensus Statement, the Utstein
template [@ringdal2008utstein]

The Karolinska trauma registry also contains a local care quality
database with specific AFs and OFIs that are determined in MMC
conferences through consensus decisions. OFIs are assessed in multiple stages,
characterised by escalating levels of scrutiny. Notably, instances of
mortality are directly referred to the multidisciplinary conference,
where, in addition to assessing OFIs, a decision is made regarding
whether the death was preventable or potentially preventable, a
classification also falling under the purview of OFIs.

From 2013 to 2017, efforts focused on identifying adverse outcomes unrelated to mortality. The review process was refined and formalized, with trauma cases initially assessed by a specialized trauma nurse for potential OFIs. In 2017, this procedure became standard practice, incorporating preliminary evaluations by specialized trauma nurses. Trauma quality data screened for AFs and cases flagged for possible care failures underwent secondary review by two specialized nurses. Potential OFIs identified in this review prompted a comprehensive evaluation of the trauma patient's case.
```{=html}
<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
```
## Participants

@@

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements
The outcome variable is an OFI, as established by the MMCs
and treated as a dichotomous variable with "Yes - At least
one OFI identified" and "No - No OFI identified".
The 11 AFs used at Karolinska University Hospital served as the exposure variable, as shown in Table 1. Each AF is labeled "original" or "manually created." Manually created AFs were calculated using trauma registry data, while original AFs were recorded in the quality database based on criteria beyond the available registry data.

The data spanning 2012 to 2022 was retrieved from the
Karolinska University Hospital trauma registry and trauma care quality
database. Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(AFs) and outcome (OFI) were retrieved from the Karolinska University
Hospital trauma care quality database. All data was anonymised to protect patient privacy.
```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias
Consensus decisions at MMC conferences may introduce misclassification bias. Additionally, AF flagging by specialized nurses is a manual process, potentially leading to selection bias. Healthcare professionals attending MMCs may not accurately recall key details impacting OFI decisions, introducing recall bias. Furthermore, professionals not present for specific cases might over- or underestimate OFI significance.

## Study size

<!-- Explain how the study size was arrived at -->

## Quantitative variables

```{=html}
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
```
## Statistical methods
The study utilized R [@R2022language] for statistical analysis. Patient characteristics between cases with and without OFI were compared using the Wilcoxon signed-rank test for continuous variables and the Chi-squared test for categorical variables.

@@

Bootstrap resampling (1,000 iterations) was applied to estimate 95% confidence intervals for sensitivity, specificity, and AUC. The significance of each AF's AUC was evaluated against chance (AUC = 0.5) and the minimum standard for diagnostic tests (AUC = 0.8) [@nahm2022receiver] using an empirical bootstrap test (1,000 iterations).

Missing values for AFs were treated as negative (i.e., "not flagged").
```{=html}
<!--

@@

<!--

@@

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

@@

The aim of this study was to evaluate the performance of individual AFs in predicting OFIs and 
to assess their suitability as a screening tool for MMCs. This study showed that no AFs was significantly successful in predicting OFIs. Four AFs ("SBP<90," "dead at 30 days," "GCS<9 and not intubated," and "CPR and thoracotomy") were equivalent to random chance in their predictive ability. The highest AUC was for ">60 minutes until first intervention,". Notably, all AFs performed significantly below the accepted standard for diagnostic tests (>0.8).

>>>>

---
title: Performance of Individual Audit Filters in Predicting Opportunities for Improvement
  in Adult Trauma Patients
author: "Josef Al-Khalili"
output:
  html_document: default
  pdf_document: default
subtitle: "A single-center registry-based retrospective cohort"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options:
  markdown:
    wrap: 72
---

```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(pROC)
library(gtsummary)
library(gt)
library(boot)
library(knitr)
```

# Abstract

@@

### Morbidity and Mortality Conferences

The MMC is a central QI technique focusing on anonymity, critical
analysis of specific adverse events, recognizing flawed approaches, and
proposing and implementing improvements [@gregor2016morbidity]. Attended
by healthcare professionals involved in trauma care, MMCs discuss themes
like unexpected mortality, morbidity, and errors
[@orlander2003morbidity], fostering error reduction and encouraging
reporting without negative feedback [@lazzara2022morbidity]. Effective
MMC implementation has been shown to increase morbidity and mortality
reporting [@kong2016analysis].

@@

We conducted a registry-based study using all trauma patients included
in both the Karolinska University Hospital trauma registry and trauma
care quality database between 2012 and 2022 to compare the performance
of AFs in their ability to predict OFI. This study was approved by
Stockholm Research Ethics Review Board, approval number 2021-02541 and
2021-03531.

@@

## Setting

Karolinska University Hospital, a Trauma level I hospital, treats around
1,500 trauma patients annually. If a patient case results in team
activation or the patient has an Injury Severity Score (ISS) of less
than 9 retrospectively, it is recorded in the Karolinska trauma
registry, which reports to SweTrau. This registry includes data on vital
signs, timings, injuries, interventions, and patient demographics as per
the European Consensus Statement (Utstein template)
[@ringdal2008utstein]

The Karolinska trauma registry also contains a local care quality
database with AFs and OFIs, identified by MMCs through consensus
decisions. OFIs are assessed in multiple stages of scrutiny. Notably,
mortality is directly referred to the MMCs in order to assess OFIs and
deciding whether the death was preventable or potentially preventable, a
classification also falling under the purview of OFIs.

From 2013 to 2017, efforts focused on
identifying adverse outcomes unrelated to mortality. Initially, trauma cases where individually assessed by a
specialized trauma nurse for potential OFIs. In 2017, this procedure
became standard practice, incorporating preliminary evaluations by
specialized trauma nurses. Cases
flagged by an AF or identified by the nurse as potential
care failures underwent a second review by two specialized nurses. Potential OFIs identified in this review
prompted review in a MMC.

```{=html}
<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
```
## Participants

@@

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements

The outcome variable is an OFI, as established by the MMCs and treated
as a dichotomous variable with "Yes - At least one OFI identified" and
"No - No OFI identified". The 11 AFs used at Karolinska University
Hospital served as the exposure variable, as shown in Table 1. Each AF
is labeled "original" or "manually created." Manually created AFs were
calculated using trauma registry data, while original AFs were recorded
in the quality database based on criteria beyond the available registry
data.

The data spanning 2012 to 2022 was retrieved from the Karolinska
University Hospital trauma registry and trauma care quality database.
Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(AFs) and outcome (OFI) were retrieved from the Karolinska University
Hospital trauma care quality database. All data was anonymised to
protect patient privacy.

```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias

Consensus decisions at MMC conferences may introduce misclassification
bias. Additionally, AF flagging by specialized nurses is a manual
process, potentially leading to selection bias. Healthcare professionals
attending MMCs may not accurately recall key details impacting OFI
decisions, introducing recall bias. Furthermore, professionals not
present for specific cases might over- or underestimate OFI
significance.

```{=html}
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
```
## Statistical methods

The study utilized R [@R2022language] for statistical analysis. Patient
characteristics between cases with and without OFI were compared using
the Wilcoxon signed-rank test for continuous variables and the
Chi-squared test for categorical variables.

@@

Bootstrap resampling (1,000 iterations) was applied to estimate 95%
confidence intervals for sensitivity, specificity, and AUC. The
significance of each AFs' AUC was evaluated against chance (AUC = 0.5)
and the minimum standard for diagnostic tests (AUC = 0.8)
[@nahm2022receiver] using an empirical bootstrap test (1,000
iterations).

Missing values for AFs were treated as negative (i.e., "not flagged").

```{=html}
<!--

@@

```{=html}
<!--

@@

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->
```
You can include code in this document like this:

@@

The aim of this study was to evaluate the performance of individual AFs
in predicting OFIs and to assess their suitability as a screening tool
for MMCs. This study showed that no AFs was significantly successful in
predicting OFIs. Four AFs ("SBP\<90," "dead at 30 days," "GCS\<9 and not
intubated," and "CPR and thoracotomy") were equivalent to random chance
in their predictive ability. The highest AUC was for "\>60 minutes until
first intervention,". Notably, all AFs performed significantly below the
accepted standard for diagnostic tests (\>0.8).
